cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Add subsections
Updated gen_expbranches

---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is where each observation is described by many features, or dimensions. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

@@

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for gen\\_multicluster().", format="latex", col.names = c("Argument", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

A branching structure captures trajectories that diverge or bifurcate from a common origin, resembling processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

@@

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in $2\text{-}D$, with optional noise dimensions to embed the structure in a $p$-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch $i$ is constructed using:

$X_1 \sim U(-2, 2)$
$X_2 = \exp(\pm s_i \cdot X_1) + \epsilon$,

where:
$\epsilon \sim U(0, 0.1)$ adds local jitter,

$s_i \in [0.5, 2]$ is randomly sampled for each branch.

The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected about $x$-axis) and even-numbered branches: growth pattern.

For $p > 2$, Gaussian noise is added to embed the $2\text{-}D$ branches in higher dimensions:

These represent irrelevant or noisy dimensions.

@@

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches in a $p$-dimensional space. The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X_1 \sim U(-2, 8)$, $X_2 = 0.5 \cdot X_1 + \epsilon$. The Branch 2 is generated from $X_1 \sim U(-6, 2)$, $X_2 = -0.5 \cdot X_1 + \epsilon$. $\epsilon \sim U(0, 0.5)$.

Branches 3 to $k$ are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

$X_1$ values span a short segment ($x_{start}$ to $x_{start} + 1$),

$X_2 = s_i \cdot (X_1 - x_{start}) + y_{start} + \epsilon$ where $s_i$ is a sampled slope from a filtered set and $\epsilon \sim U(0, 0.2)$.

This setup yields $k$ spatially distinct, noisy linear trajectories.

For $p > 2$, the remaining dimensions are pure noise:

$$
X_j \sim N(0, 0.05^2)\text{ for }j = 3, \dots, p.
$$

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is where each observation is described by many features, or dimensions. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

@@

```{r main-tb-pdf, eval=knitr::is_latex_output()}
main_tb |> 
  kable(caption = "The main arguments for gen\\_multicluster().", format="latex", col.names = c("Argument", "Explanation"), booktabs = T, table.pos = "H")  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

A branching structure captures trajectories that diverge or bifurcate from a common origin, similar processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

@@

#### `gen_expbranches()`

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in $2\text{-}D$, with optional noise dimensions to embed the structure in $p\text{-}D$. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch $i$ is constructed using $X_1 \sim U(-2, 2)$ and $X_2 = \exp(\pm s_iX_1) + \epsilon$, where: $\epsilon \sim U(0, 0.1)$ adds local jitter, $s_i \in [0.5, 2]$ is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected) and even-numbered branches: growth pattern. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

#### `gen_linearbranches()`

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches in $p\text{-}D$. The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches $1$ and $2$ are initialized with fixed slopes and intercepts. The Branch $1$ is generated from $X_1 \sim U(-2, 8)$, $X_2 = 0.5X_1 + \epsilon$, where $\epsilon \sim U(0, 0.5)$. The Branch $2$ is generated from $X_1 \sim U(-6, 2)$, $X_2 = -0.5X_1 + \epsilon$, where $\epsilon \sim U(0, 0.5)$. Branches $3$ to $k$ are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches. $X_1$ values span a short segment ($x_{start}$ to $x_{start} + 1$) and $X_2 = s_i(X_1 - x_{start}) + y_{start} + \epsilon$ where $s_i$ is a sampled slope from a filtered set and $\epsilon \sim U(0, 0.2)$. This setup yields $k$ spatially distinct, noisy linear trajectories.

For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

#### `gen_curvybranches()`

@@

#### `gen_orglinearbranches()`

@@

#### `gen_orgcurvybranches()`

@@

#### `gen_gridcube()`

@@

#### `gen_unifcube()`

@@

#### `gen_cubehole()`

@@

#### `gen_quadratic()`

@@

#### `gen_cubic()`

@@

#### `gen_pyrrect()`

@@

#### `gen_pyrtri()`

@@

#### `gen_pyrstar()`

@@

#### `gen_pyrholes()`

@@

#### `gen_scurve()`

@@

#### `gen_scurvehole()`

@@

#### `gen_circle()`

@@

#### `gen_curvycycle()`

@@

#### `gen_unifsphere()`

@@

#### `gen_gridedsphere()`

@@

#### `gen_clusteredspheres()`

@@

#### `gen_hemisphere()`

@@

#### `gen_crescent()`

@@

#### `gen_curvycylinder()`

@@

#### `gen_sphericalspiral()`

@@

#### `gen_helicalspiral()`

@@

#### `gen_conicspiral()`

@@

#### `gen_nonlinear()`

### 2025-06-24 paper-cardinalR.tex
Updated gen_expbranches

A branching structure captures trajectories that diverge or bifurcate from a common origin, resembling processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

@@

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in \(2\text{-}D\), with optional noise dimensions to embed the structure in a \(p\)-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch \(i\) is constructed using:

\(X_1 \sim U(-2, 2)\)
\(X_2 = \exp(\pm s_i \cdot X_1) + \epsilon\),

where:
\(\epsilon \sim U(0, 0.1)\) adds local jitter,

\(s_i \in [0.5, 2]\) is randomly sampled for each branch.

The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected about \(x\)-axis) and even-numbered branches: growth pattern.

For \(p > 2\), Gaussian noise is added to embed the \(2\text{-}D\) branches in higher dimensions:

These represent irrelevant or noisy dimensions.

@@

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches in a \(p\)-dimensional space. The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from \(X_1 \sim U(-2, 8)\), \(X_2 = 0.5 \cdot X_1 + \epsilon\). The Branch 2 is generated from \(X_1 \sim U(-6, 2)\), \(X_2 = -0.5 \cdot X_1 + \epsilon\). \(\epsilon \sim U(0, 0.5)\).

Branches 3 to \(k\) are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

\(X_1\) values span a short segment (\(x_{start}\) to \(x_{start} + 1\)),

\(X_2 = s_i \cdot (X_1 - x_{start}) + y_{start} + \epsilon\) where \(s_i\) is a sampled slope from a filtered set and \(\epsilon \sim U(0, 0.2)\).

This setup yields \(k\) spatially distinct, noisy linear trajectories.

For \(p > 2\), the remaining dimensions are pure noise:

\[
X_j \sim N(0, 0.05^2)\text{ for }j = 3, \dots, p.
\]

>>>>

A branching structure captures trajectories that diverge or bifurcate from a common origin, similar processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table \ref{tab:branching-tb-pdf} outlines these functions. The main arguments of the functions described in Table \ref{tab:arg-branching-tb-pdf}.

@@

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in \(2\text{-}D\), with optional noise dimensions to embed the structure in \(p\text{-}D\). These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch \(i\) is constructed using \(X_1 \sim U(-2, 2)\) and \(X_2 = \exp(\pm s_iX_1) + \epsilon\), where: \(\epsilon \sim U(0, 0.1)\) adds local jitter, \(s_i \in [0.5, 2]\) is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected) and even-numbered branches: growth pattern. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches in \(p\text{-}D\). The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts. The Branch \(1\) is generated from \(X_1 \sim U(-2, 8)\), \(X_2 = 0.5X_1 + \epsilon\), where \(\epsilon \sim U(0, 0.5)\). The Branch \(2\) is generated from \(X_1 \sim U(-6, 2)\), \(X_2 = -0.5X_1 + \epsilon\), where \(\epsilon \sim U(0, 0.5)\). Branches \(3\) to \(k\) are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches. \(X_1\) values span a short segment (\(x_{start}\) to \(x_{start} + 1\)) and \(X_2 = s_i(X_1 - x_{start}) + y_{start} + \epsilon\) where \(s_i\) is a sampled slope from a filtered set and \(\epsilon \sim U(0, 0.2)\). This setup yields \(k\) spatially distinct, noisy linear trajectories.

For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).
